---
title: "Intro to Data Science with R"
subtitle: "Day 1"
author: "Econ 122"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
---

## Course Overview & Foundations

Welcome to the course! Today, we will lay the groundwork for our journey into data science using R.

-   Course Overview
-   The R Environment & RStudio
-   Basic R Syntax & Data Types
-   Introduction to R Markdown

## Course overview {.smaller}

- See the [Syllabus](https://github.com/econ122-f25/home/blob/main/ECON122_F2025_DataScience_StatisticalLearning.pdf) for complete breakdown

- Typical class:
    + Short lecture on the concept we will focus on
    + In class exercise to apply the concepts we just covered

- Evaluation:
    + Class attendance/Participation [5%]
    + Problem sets (0/1/2 scale) [15%]
    + Team projects              [15%]
    + 2 midterm exams            [40%]
    + Final project              [25%]

- Note that 40% of your grade is from exams where you have to answer questions and write code in real time
- Feel free to use GenAI but you won't be able to use it on the exams

## How to succeed

- Stay up to date with the course outline (I will update regularly)
- Come to class prepared and on time
- Ask questions if you start getting lost
    + In class
    + Tutor OH
    + My OH

## What is Data Science?

Data science is an interdisciplinary field that uses scientific methods, processes, algorithms, and systems to extract knowledge and insights from structured and unstructured data.

### The "Data Science Lifecycle"

-   **Ask:** Define the problem and form a hypothesis.
-   **Wrangle:** Get, clean, and prepare data.
-   **Explore:** Conduct exploratory data analysis (EDA) and visualize the data.
-   **Model:** Build predictive or descriptive models.
-   **Communicate:** Share findings and insights.

Roles used to be split but now merged

## Why R?

R is a powerful, open-source programming language and environment specifically designed for statistical computing and graphics. 

-   **Rich Ecosystem:** Thousands of packages for every data task imaginable.
-   **Visualization:** Unrivaled capabilities with packages like `ggplot2`.
-   **Statistical Power:** Built by statisticians, for statisticians.
-   **Reproducibility:** Excellent tools for creating dynamic reports and dashboards.

- **Python** is also commonly used. It is more general use than **R**.

## Introducing RStudio

RStudio is an integrated development environment (IDE) that makes working with R much easier and more productive. It consists of four main panes:

-   **Console:** Where you run R code directly.
-   **Source:** Where you write and save your R scripts and R Markdown files.
-   **Environment/History:** Where you see the variables and functions you have created.
-   **Files/Plots/Packages/Help/Viewer:** A multi-tab pane for file management, viewing plots, managing packages, and more.



## Basic R Syntax: Assignment

Let's start with the basics. R's syntax is simple and intuitive for a programming language.

### Assignment Operator

Use `<-` to assign a value to a variable. The `=` sign also works, but `<-` is the conventional R style.

```{r, eval=TRUE}
# Assign a value to x
x <- 10

# Print the value of x
print(x)
```



## Live Example: Assignment

Let's see this in action. When you "knit" this document, the code below will be executed and the output will be displayed.

```{r}
# Assign a value to a variable
my_number <- 42
my_text <- "Hello, R!"

# Print the variables
paste("my_number is:", my_number)
paste("my_text is:", my_text)
```



## Basic arithmetic

You can use R for basic arithmetic operations.

-   Addition: `+`
-   Subtraction: `-`
-   Multiplication: `*`
-   Division: `/`
-   Exponentiation: `^` or `**`

```{r, eval=TRUE}
5 * 3 + 2
```


## Data Types: Numbers

The two main numeric data types are integers and doubles (real numbers).

-   **Integer:** Whole numbers, specified with `L` (e.g., `10L`).
-   **Double (Numeric):** Decimal numbers, the default numeric type.
-  **Note:** `class()` allows you to see the data type of your variable.

```{r, eval=TRUE}
class(10)
class(10L)
```

## Data Types: Strings (Character)

Text data is stored as a character string. You can use either single quotes (`'`) or double quotes (`"`).

```{r, eval=TRUE}
my_string <- "Data Science is fun!"
print(my_string)
```

## Data Types: Logical (Boolean)

Logical values are either `TRUE` or `FALSE`. You can also use `T` and `F` as shorthand.

```{r, eval=TRUE}
5 > 3
5 == 5
```


## Introduction to R Markdown

R Markdown is a powerful tool for creating dynamic, reproducible documents that combine R code and its output with narrative text.

-   **Narrative Text:** Use Markdown syntax for headings, lists, and bold text.
-   **Code Chunks:** Write R code in special blocks called "chunks."
-   **Output:** The results of your code, including tables and plots, are automatically embedded.



## The Structure of an R Markdown File

An R Markdown file (`.Rmd`) has three main components:

### 1. The YAML Header

```yaml

title: "My First R Markdown Document"
author: "Your Name"
date: "2025-08-04"
output: html_document

```
This specifies document metadata and the output format.



## R Markdown: Narrative & Code

### 2. The Narrative Text

This is just regular text using Markdown syntax.

`# A main heading`
`This is some **bold** text.`

### 3. The Code Chunks

Code chunks are enclosed in backticks and curly braces:

````markdown
```r
# Your R code goes here
summary(cars)
````



## Live Example: R Markdown Concept

When you knit an R Markdown file, the code is executed and the output is inserted into the final document.

```{r}
# Imagine this is an R Markdown chunk
my_vector <- c(1, 2, 3, 4, 5)
mean(my_vector)
```


## The "Knit" Button

In RStudio, the "Knit" button is what brings it all together. It executes your R code and renders the final document to your desired format (e.g., HTML, PDF, Word).

-   It's the key to **reproducibility**.
-   It ensures your analysis and the report are always in sync.


## Summary and Next Steps

### Today's Summary

-   We discussed what **Data Science** is.
-   We introduced **R** and **RStudio**.
-   We learned **basic R syntax** for assignment and arithmetic.
-   We explored fundamental **data types**: numbers, strings, and logicals.
-   We had a first look at the power of **R Markdown**.

### Looking Ahead to Day 2

- Next class, we'll dive deeper into R Markdown and introduce **version control** with Git and GitHub, which are essential for collaboration.

## For rest of class

- Set up R, Rstudio

- Work on the `day1_activity`

- Set up Git, GitHub

- Ask me questions if you run into issues
  

