---
title: "Day 4: In-Class Activity - Student Data Wrangling"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

## Welcome to Your In-Class Activity!

This activity will give you hands-on practice with the data reshaping and joining techniques we discussed today. You'll work with simulated student and course data from the Claremont Colleges.

**Estimated Time:** 40 minutes

**Goal:** By the end of this activity, you should be comfortable applying `tidyr` functions for reshaping and `dplyr` functions for joining datasets.

## Part 1: Data Setup (5 minutes)

First, let's create the datasets you'll be working with. Run the following R code chunk to create three tibbles: `students_raw`, `colleges_info`, and `course_enrollment`.

```{r data_setup, echo=TRUE}
# Raw student data (untidy for reshaping)
students_raw <- tibble(
  student_id = c(101, 102, 103, 104, 105, 106),
  name = c("Alice", "Bob", "Charlie", "David", "Eve", "Frank"),
  Pomona = c("Yes", "No", "Yes", "No", "No", "No"),
  Scripps = c("No", "Yes", "No", "No", "Yes", "No"),
  CMC = c("No", "No", "No", "Yes", "No", "No")
)

# Information about the colleges
colleges_info <- tibble(
  college_name = c("Pomona", "Scripps", "Claremont McKenna", "Harvey Mudd", "Pitzer", "Keck Graduate Institute"),
  college_id = c(1, 2, 3, 4, 5, 6),
  type = c("Liberal Arts", "Liberal Arts", "Liberal Arts", "STEM", "Liberal Arts", "Graduate")
)

# Course enrollment data
course_enrollment <- tibble(
  student_id = c(101, 101, 102, 103, 104, 105, 107),
  course_code = c("DS101", "MATH200", "BIO101", "DS101", "ECON300", "CHEM101", "CS100"),
  semester = c("Fall23", "Fall23", "Spring24", "Fall23", "Spring24", "Fall23", "Fall23")
)

# Display the raw dataframes
print(students_raw)
print(colleges_info)
print(course_enrollment)
```

## Part 2: Reshaping Data (10 minutes)

The `students_raw` data is in a **wide** format, where each college is its own column. Your task is to transform it into a **tidy (long)** format, where you have one column for `college_name` and another for `is_enrolled` (indicating "Yes" or "No").

**Task:** Use `pivot_longer()` to reshape `students_raw` into a new tibble called `students_tidy`.

```{r reshape_task, echo=TRUE}
# Your code here: Reshape students_raw into students_tidy


# Display the tidy student data


```

**Hint:** Remember to specify which columns to pivot, what to name the new "names" column, and what to name the new "values" column. You might also want to filter out the "No" entries after pivoting to only show the colleges a student *is* enrolled in.

## Part 3: Mutating Joins (15 minutes)

Now that you have `students_tidy` and `colleges_info`, let's combine them to get more details about the colleges.

**Task 3.1: Add College Type**
Use a mutating join to add the `type` of college (e.g., "Liberal Arts", "STEM") from `colleges_info` to your `students_tidy` data. Store the result in `students_with_college_type`.

```{r mutate_join_1, echo=TRUE}
# Your code here: Join students_tidy with colleges_info


# Display the result


```

**Task 3.2: Combine with Course Enrollment**
Now, let's combine `students_with_college_type` with `course_enrollment` to see which courses each student is taking. Store the result in `full_student_data`.

```{r mutate_join_2, echo=TRUE}
# Your code here: Join students_with_college_type with course_enrollment


# Display the result


```

**Hint:** Pay attention to the common columns between the dataframes for the `by` argument in your join functions. Consider which type of mutating join (`inner`, `left`, `full`) makes the most sense for each task to ensure you keep the desired rows.

## Part 4: Filtering Joins (10 minutes)

Filtering joins are great for identifying subsets of your data.

**Task 4.1: Find Enrolled Students with Courses**
Use a filtering join to identify which students from `students_raw` (before tidying) are actually enrolled in a course listed in `course_enrollment`. This should *only* return columns from `students_raw`.

```{r filter_join_1, echo=TRUE}
# Your code here:


# Display the result


```

**Task 4.2: Find Students Without Courses**
Now, use another filtering join to identify which students from `students_raw` are *not* currently enrolled in any course listed in `course_enrollment`.

```{r filter_join_2, echo=TRUE}
# Your code here:


# Display the result


```


## Reflection & Discussion

* What was the most challenging part of this activity for you?
* Can you think of other real-world scenarios where reshaping data would be necessary?
* When might an `inner_join()` be more appropriate than a `left_join()`?
* How could `anti_join()` be useful for data cleaning or identifying anomalies in a larger dataset?
