---
title: "Day 6: Advanced EDA & Patterns - In-Class Activity"
author: "Econ 122"
date: "`r Sys.Date()`"
output:
  html_document:
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, out.width = '70%', fig.align = 'center') # Adjusted out.width for document readability

# Load all necessary libraries
library(ggplot2)
library(dplyr)
library(reshape2) # For melt function
library(readr) # For reading CSV files

#Download happiness data and clean it
whr_url <- "https://raw.githubusercontent.com/mgelman/data/master/whr2023.csv"
whr_data <- read_csv(whr_url)

whr_data <- whr_data %>%
  rename(
    happiness = `Happiness score`, 
    Logged_GDP_per_capita = `Logged GDP per capita`,
    Social_support = `Social support`,
    Healthy_life_expectancy = `Healthy life expectancy`, 
    Freedom_to_make_life_choices = `Freedom to make life choices`,
    Generosity = `Generosity`,
    Perceptions_of_corruption = `Perceptions of corruption`,
    Regional_indicator = `Regional indicator`,
    Country_name = `Country name`
  ) %>%
  na.omit() 

# Generate raw GDP per capita from the logged version for Task 3.1
whr_data <- whr_data %>%
  mutate(GDP_per_capita_raw = exp(Logged_GDP_per_capita))

# Create happiness bins for stacked bar chart (Task 3.2)
whr_data <- whr_data %>%
  mutate(
    happiness_bin = cut(happiness,
                        breaks = c(0, 4.5, 6.5, Inf), # Example bins: Low, Medium, High
                        labels = c("Low Happiness", "Medium Happiness", "High Happiness"),
                        right = FALSE) # Include lower bound, exclude upper
  )

```

# 📊 Day 6: Advanced EDA & Patterns - In-Class Activity

Welcome to your in-class activity for Day 6! Today, you'll apply the advanced Exploratory Data Analysis (EDA) techniques we discussed, focusing on identifying patterns, relationships, anomalies, and data transformations, using an **actual World Happiness Report dataset (2023)**. This dataset tracks **happiness scores**, **GDP per capita**, **social support**, and **healthy life expectancy** across various countries and regions. It offers **interesting variations** and **distinct patterns** throughout the world, making it ideal for deeper exploration of global well-being and economic trends.

**Estimated Time:** 40 minutes

**Instructions:**

1.  Work through each section.
2.  Execute the R code chunks as you go.
3.  Answer any questions or complete any coding tasks.
4.  Feel free to discuss with your peers!

---

## Part 1: Global Happiness & Economy Relationships (15 minutes)

In this section, you'll explore relationships between happiness scores and economic indicators using scatter plots, faceting, and correlation matrices.

### Task 1.0:  GDP Per Capita vs. Happiness Score with Trend Line
Create a scatter plot showing `GDP_per_capita_raw` (GDP per capita) on the x-axis and `happiness` (Happiness Score) on the y-axis from the `whr_data` dataset. Add a linear regression line (`geom_smooth(method = "lm", se = FALSE)`) to visualize the overall trend.

```{r task1.0}
# Your code here for GDP vs Happiness Score with linear trend

```

### Task 1.1: Log GDP Per Capita vs. Happiness Score with Trend Line
Do the same thing but use  `Logged_GDP_per_capita` (Log GDP per capita) on the x-axis instead. What are the main differences between the `raw` and `log` versions?

```{r task1.1}
# Your code here for Log GDP vs Happiness Score with linear trend

```

### Task 1.2: Faceting by Regional Indicator

Take the previous plot (Log GDP Per Capita vs. Happiness Score) and use `facet_wrap` on `Regional_indicator` to create separate plots for each region. Observe how the relationship between Log GDP and happiness changes across these different regions. Do you notice **more distinct patterns** or different slopes for certain regions, indicating varying factors influencing happiness?

```{r task1.2}
# Your code here for faceted Log GDP vs Happiness Score plot by Regional Indicator

```

### Task 1.3: Visualizing a Correlation Matrix

Generate a heatmap of the correlation matrix for the numerical measurements in `whr_data` (`happiness`, `Logged_GDP_per_capita`, `Social_support`, `Healthy_life_expectancy`, `Freedom_to_make_life_choices`, `Generosity`, `Perceptions_of_corruption`). This will help you quickly identify strong positive and negative linear relationships between these global well-being indicators.

```{r task1.3}
# Your code here for correlation heatmap

```

---

## Part 2: Anomaly Detection in Global Data (15 minutes)

In this section, you'll use visual and statistical methods to identify potential anomalies (outliers) in the `happiness` (happiness score) variables of the `whr_data` dataset. These might represent countries with unusually high/low happiness for their region/economic context.

### Task 2.1: Outliers in Happiness Score

Explore outliers in the `happiness` score using both visual and statistical methods.

#### Task 2.1.1: Box Plot for Happiness Score Outliers

Create a box plot of `happiness` to visually identify outliers. Given the nature of real-world data, you might observe some interesting outliers.

```{r task2.1.1}
# Your code here for box plot of happiness

```

#### Task 2.1.2: IQR Method for Happiness Score

Calculate the IQR for `happiness`. Identify and print the `Country_name`, `Regional_indicator`, and `happiness` for any entries that fall outside the $Q1 - 1.5 \\times IQR$ and $Q3 + 1.5 \\times IQR$ bounds. These could be countries with unusually low or high happiness scores.

```{r task2.1.2}
# Your code here to identify and print IQR outliers

```

#### Task 2.1.3: Z-score Method for Happiness Score (Identifying More Outliers)

Calculate the Z-score for each country's `happiness`. Identify and print the `Country_name`, `Regional_indicator`, and `happiness` for any entries with an absolute Z-score greater than **1.5**. Using this lower threshold will typically identify **even more outliers**, allowing for a broader examination of unusual data points.

```{r task2.1.3}
# Your code here to identify and print Z-score outliers using a lower threshold

```

---

## Part 3: Data Transformation & Categorical Visualization (10 minutes)

This section focuses on transforming data to improve visualization and exploring relationships between categorical variables.

### Task 3.1: Effect of Log Transformation on GDP Per Capita

Explore the distribution of raw GDP per capita and its log-transformed version. Create two histograms side-by-side: one for `GDP_per_capita_raw` and one for `Logged_GDP_per_capita`. Observe how the log transformation changes the distribution's shape, making it more symmetric and suitable for linear modeling.

```{r task3.1}
# Your code here for histograms of original and log-transformed GDP per capita

```

### Task 3.2: Stacked Bar Chart of Happiness Distribution by Region

Create a stacked bar chart showing the **count of countries** within each `Regional_indicator` that fall into different `happiness_bin` categories ("Low Happiness", "Medium Happiness", "High Happiness"). This will allow you to visualize the distribution of happiness levels across regions. What do you notice?

```{r task3.2}
# Your code here for stacked bar chart of happiness distribution by region

```

---

## ✨ Activity Wrap-up

Great job working through these advanced EDA concepts! You've practiced:

* Using **faceting** to compare relationships across groups (specifically by `Regional_indicator`, revealing distinct patterns related to happiness and economic indicators).
* Visualizing trends with **`geom_smooth()`**.
* Identifying linear relationships with **correlation matrices** and **heatmaps**.
* Detecting anomalies using **box plots**, **Z-score**, and **IQR methods**, including identifying **extreme outliers** in happiness scores.
* Transforming skewed data (like GDP per capita) with **log transformations** to improve visualization.
* Visualizing relationships between categorical variables using **stacked bar charts** to show distributions.

### Discussion Points:

* Compare the Z-score and IQR methods for detecting outliers in `happiness`. Which method identified more outliers and why might that be the case for these specific variables?
* How did faceting by `Regional_indicator` provide different insights into the relationship between Log GDP per capita and Happiness Score, potentially highlighting socio-economic disparities?
* Think of another real-world dataset where identifying obvious data errors or extreme outliers (like a country with an unusually low life expectancy given its high GDP) would be crucial before further analysis related to global well-being.

