---
title: "Day 7 Activity: Data Cleaning Challenge"
author: "Student Name"
date: "`r Sys.Date()`"
output:
  html_document:
---

# 🔍 In-Class Activity: The Messy Sales Dataset

**Estimated Time:** 40 minutes

**Goal:** In this activity, you will act as a junior data analyst tasked with cleaning and preparing a new sales dataset. This dataset is "fresh off the press" and contains common data issues like inconsistent date formats, messy text, and missing values.

Your mission is to use the skills you learned today to get the data into a clean, tidy format suitable for analysis.

## Setup

First, let's load the necessary packages and create our messy dataset.

```{r setup, message=FALSE, warning=FALSE}
# Load the required packages
library(tidyverse)
library(lubridate)
library(stringr)
library(naniar)

# Create the messy dataset
sales_data <- tibble(
  order_id = c("ORD-20230115", "ORD-20230220", "ORD-20230325", "ORD-20240410", "ORD-20230505", "ORD-20230630", "ORD-20230712", "ORD-20220818", "ORD-20230920", "ORD-20231105"),
  order_date_str = c("Jan 15, 2023", "20/02/2023", "2023-03-25", "April 10 2024", "05-May-2023", "2023-06-30", "July 12, 2023", "18-Aug-2022", "Sep 20, 2023", "Nov 5, 2023"),
  shipping_time_str = c("10:30 AM", "14:05 PM", "12:00 PM", "9:15 AM", "11:45 AM", NA, "13:30 PM", "10:00 AM", "15:30 PM", "8:45 AM"),
  customer_info = c("John.D-New", "Jane_S_Old", "Peter_J-New", "Maria.L-New", "Mike_A_Old", "Sarah.W-New", "Sam.T_Old", "Emily_R_New", "David_K_New", "Lisa.M-Old"),
  customer_segment = c("Tier 1", "Tier 2", "Tier 1", "Tier 1", "Tier 2", "Tier 1", "Tier 2", "Tier 1", "Tier 1", "Tier 2"),
  product_revenue = c(150.75, 200.50, NA, 85.25, -50.00, 45.99, NA, 1250.00, 320.15, 89.99),
  feedback_status = c("Complete", "Pending", "N/A", "Complete", "Not provided", "Pending", "N/A", "Complete", "Pending", "Complete")
)

# Look at the data
sales_data
```

---

## Part 1: Dates and Times (15 minutes)

Your first task is to fix the `order_date_str` and `shipping_time_str` columns.

### **Step 1: Fix the Order Date**

The `order_date_str` column contains dates in various formats. Convert this column to a proper `Date` object using `lubridate`.

**Hint:** You will need to use a combination of `mutate()` and `parse_date_time()`.

```{r part1_step1}
# Your code goes here
# sales_data %>%
#   mutate(clean_date = ...)
```

### **Step 2: Combine and Calculate**

The `shipping_time_str` is a character column. Convert it to a time format, then combine it with your clean `order_date` to create a single `POSIXct` timestamp. Finally, calculate the duration between the first and last order.

**Hint:** Use `paste0()` to combine the date and time strings before using `lubridate`'s parsing functions. Use `difftime()` for the calculation.

```{r part1_step2}
# Your code goes here
```

---

## Part 2: String Manipulation (15 minutes)

The `customer_info` column is a mess! It contains the customer's first name, last initial, and a status (either "New" or "Old"). You need to extract these pieces of information into separate columns.

### **Step 3: Extract Customer Status**

Extract the customer status ("New" or "Old") into a new column called `customer_status`.

**Hint:** Use `str_extract()` with a simple pattern.

```{r part2_step3}
# Your code goes here
```

### **Step 4: Clean Customer Names**

Clean the `customer_info` column to create a new column called `customer_name` that contains only the first name and last initial (e.g., "John D"). Remove any special characters (`.` or `_`).

**Hint:** Use `str_replace_all()` to remove the unwanted characters and `str_extract()` to pull out the first name and initial.

```{r part2_step4}
# Your code goes here
```

---

## Part 3: Handling Missing Values & Imputation (10 minutes)

Now, let's address the missing data in our dataset. This is a crucial step for accurate analysis.

### **Step 5: Standardize and Impute**

The `feedback_status` column uses both "N/A" and "Not provided" as missing value indicators. First, standardize these by converting both to `NA`. Then, identify the missing value in `product_revenue` and **impute** it by replacing it with the mean of the other values **within its `customer_segment` group**.

**Hint:** Use `na_if()` to handle the string-based missing values. For imputation, you can use `group_by()` and `mutate()` together.

```{r part3_step5}
# Your code goes here
```

### **Step 6: Visualize the Clean Data**

Once you've standardized and imputed your data, visualize the missing values one last time with `vis_miss()` to confirm that all of your intended changes have been successful.

```{r part3_step6}
# Your code goes here
```

---

# 🏆 Challenge Activities

If you finish early or want extra practice, try these additional challenges that build on today's topics!

## Challenge 1: Advanced Date Operations (10 minutes)

Create some additional time-based variables from your cleaned data.

### **Challenge 1A: Time Periods**

Calculate how many days have passed between each order and today's date. Also, determine which day of the week each order was placed and which quarter of the year.

**Hint:** Use `today()`, `wday()`, and `quarter()` functions.

```{r challenge1a}
# Your code goes here
```

### **Challenge 1B: Seasonal Analysis**

Create a new variable called `season` based on the order month:
- Winter: Dec, Jan, Feb  
- Spring: Mar, Apr, May
- Summer: Jun, Jul, Aug  
- Fall: Sep, Oct, Nov

**Hint:** Use `month()` and `case_when()` to create the seasonal categories.

```{r challenge1b}
# Your code goes here
```

## Challenge 2: Advanced String Manipulation (10 minutes)

Let's do some more complex text processing with the order IDs and customer information.

### **Challenge 2A: Extract Order Numbers**

Extract just the numeric portion from each `order_id` (e.g., from "ORD-20230115" extract "20230115") and create a new column called `order_number`.

**Hint:** Use `str_extract()` with the regex pattern `\\d+` to extract all digits.

```{r challenge2a}
# Your code goes here
```

### **Challenge 2B: Customer Email Generator**

Create mock email addresses for each customer using their cleaned name. The format should be: `firstname.lastinitial@company.com` (all lowercase).

**Hint:** Use `str_to_lower()`, `str_replace()`, and `str_c()` to build the email addresses.

```{r challenge2b}
# Your code goes here
```

### **Challenge 2C: Order ID Validation**

Check if all order IDs follow the correct format: "ORD-" followed by exactly 8 digits. Create a logical column called `valid_order_id` that indicates whether each ID is properly formatted.

**Hint:** Use `str_detect()` with the pattern `^ORD-\\d{8}$` where `^` means start of string, `\\d{8}` means exactly 8 digits, and `$` means end of string.

```{r challenge2c}
# Your code goes here
```

## Challenge 3: Missing Data Patterns (10 minutes)

Explore missing data patterns more deeply.

### **Challenge 3A: Missing Data Summary**

Create a summary table showing the percentage of missing values for each column in your dataset.

**Hint:** You can use `miss_var_summary()` from the `naniar` package.

```{r challenge3a}
# Your code goes here
```

### **Challenge 3B: Conditional Missing Values**

Create a scenario where you introduce some additional missing values based on conditions. For example, set `product_revenue` to `NA` for all "Tier 2" customers with revenue less than $100. Then visualize the new missing pattern.

**Hint:** Use `if_else()` or `case_when()` within `mutate()` to create conditional missingness.

```{r challenge3b}
# Your code goes here
```

## Challenge 4: Data Quality Checks (5 minutes)

### **Challenge 4A: Revenue Validation**

Check for any unusual values in the `product_revenue` column. Are there any negative values or values that seem unreasonably high? Create a summary of revenue statistics by customer segment.

**Hint:** Use `summary()`, `group_by()`, and `summarise()` to explore the revenue distribution.

```{r challenge4a}
# Your code goes here
```

### **Challenge 4B: Date Consistency Check**

Verify that all order dates fall within a reasonable range (e.g., within 2023). Create a logical check to identify any dates that fall outside the expected year.

**Hint:** Use `year()` and logical operators to check date consistency.

```{r challenge4b}
# Your code goes here
```

---

## 🎯 Activity Wrap-up

Congratulations on completing the data cleaning challenge! You have successfully practiced handling three of the most common data wrangling tasks:

1.  **Date/Time Manipulation:** Converting messy date strings into usable formats.
2.  **String Manipulation:** Extracting and cleaning specific information from text.
3.  **Missing Value Handling:** Identifying, standardizing, and imputing missing data.

### **Discussion:**

Once everyone has finished, let's discuss your findings. Be prepared to answer the following questions:

* Which part of the activity was the most challenging for you?
* What was the most surprising thing you learned about working with messy data?
* If this were a real dataset, what would be your next steps for analysis?
* For those who completed the challenge activities: What additional insights did you gain about the data quality?